Neighborhoods and Felony Disenfranchisement: The Case of New York City

### 2019-05-17 markdown/kmorris_thesis.Rmd
clean up maps

---
title: "Rethinking Felony Disenfranchisement: The Case of New York City and Miami"
author: "Kevin Morris"
date: "April 22, 2019"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
bibliography: "../research/citations.bib"
link-citations: yes
---

@@

Felony disenfranchisement has received much attention over the past few years. This is true in both the academic world (where researchers have dug into the political ramifications in such races as the 2000 presidential election) and in the political world as states such as Florida and Louisiana have gradually moved to dismantle their systems of disenfranchisement. Throughout this process, however, many researchers have focused on the *individual* effects of felony disenfranchisement. They have examined the effect of disenfranchisement on the political participation of individuals, including after they are no longer disenfranchised, but have generally neglected to examine the intersection of spatially concentrated policing and incarceration patterns with the political effects of disenfranchisement. This paper seeks to increase our understanding of the spatial implications of felony disenfranchisement by examining the character of disenfranchisement in New York City and Miami-Dade County in Florida.

@@

Of course, despite dire over-incarceration in some neighborhoods, the number of incarcerated individuals is relatively low compared to the number of voters. In New York, for instance, `r comma(readRDS("./temp/number_incarcerated.rds"))` individuals were imprisoned in New York State in early 2019, compared with XX actively registered voters. Despite the lower share of residents who are disenfranchised through imprisonment, there is reason to believe the policy impacts more individuals than just those imprisoned. As discussed above, previous research has demonstrated that felony disenfranchisement reduces turnout even among Black voters whose rights are not stripped. This research has found, in particular, that eligible Black voters are less likely to cast a ballot in states where felony disenfranchisement policies are harsher. Understanding the distribution of this effect is important for understanding its political implications. Are voters who live in the same neighborhood as the disenfranchised less likely to cast a ballot because their neighbors have been removed from the community? Or does felony disenfranchisement operate cast its chilling effect on turnout more broadly, crossing even into neighborhoods where voters have not been removed from the community?

@@

The third dataset used to estimate the extent to which a neighborhood is directly impacted by felony disenfranchisement policies in New York State comes from another freedom of information request. The individual-level incarceration data referenced above does not include information about the neighborhood an incarcerated person lived in prior to their conviction. Although the State of New York does not make this information available at the individual level, the author has received prison-admission counts by zip-code for the years XX - XX.

@@

## Identification of Lost Voters {-}
In this analysis, I offer a slightly different defintion of "lost voter" than much of the literature. Many recent papers have attempted to identify relationships between the number of *potentially* lost voters and turnout. Here, I explore whether the disenfranchisement of voters with a history of participating in elections is related to neighborhood turnout. In this analysis, "lost voters" are individuals ineligible to cast a ballot on a given election day who have cast a ballot in the previous ten years. To study the effect of felony disenfranchisement on voting at the local level (and its potential implications for the distribution of political power at the local level), I use turnout in the most recent election for city-wide office in New York - the mayoral election which took place on November 7^th^, 2017. Lost voters, therefore, are all individuals who were incarcerated or on parole on November 7^th^, 2017, and had cast a ballot between 2007 and 2016.

@@

After matching neighborhoods, I use a simple regression to test whether turnout in the 2017 mayoral election was different in areas with lost voters. Census tract and block group level turnout rates are calculated using the geocoded voter file. Each voter's record indicates whether the voter participated in the 2017 general election,^[The New York State voter file is not without errors. For instance, the voter file erroneously does not record the author as participating in the 2017 mayoral election. However, insofar as these errors are not geographically correlated, there is little reason to believe they impact the results of this analysis. Unfortunately, testing the distribution of these errors is not possible.] which are then aggregated to estimate the number of ballots cast in each neighborhood. The number of ballots cast is divided the voting age population.

@@

Matching methodologies, of course, only allow us to test the effect of being treated - here, losing a voter for the 2017 election. The model above does not allow for different effects on turnout based on *how many* voters a neighborhood lost. In Table \ref{tab:trad-reg} below, I adopt a standard ordinary least squares regression to investigate whether lost voters are associated with lower turnout rates in the 2017 election. This regression uses the same covariates that were used in the matching procedure described above. Each neighborhood is weighted by its voting age population, and robust standard errors are clustered by city council district.

@@

stargazer(bg_model, bg_model2, tract_model, tract_model2,
          column.separate = c(2, 2),
          column.labels = c("Block Group Level", "Tract Level"),
          header = F,
          type = "latex",
          se = list(bg_model_ses, bg_model2_ses, tract_model_ses, tract_model2_ses),
          covariate.labels = c("Lost Voters", "Lost Voters X Share Non-Hispanic Black", "Median Income (Thousands of Dollars)", "Percent Latino", "Percent Non-Hispanic Black", "Percent Non-Hispanic White",
                               "Percent With Some College", "Median Age", "Registration Rate", "Percent Democrats", "Percent Noncitizen", "Percent Won by City Council Representative"),
          dep.var.labels = c("Turnout Rate"),
          title = "(\\#tab:trad-reg) Standard Regression",
          table.placement = "H",
          style = "ajps",
          omit.stat = c("f", "ser"),
          notes = "Robust standard errors (clustered by city council district) in parentheses.",
          omit = "boro")
```
The results presented in Table \ref{tab:trad-reg} align very closely with the estimated effect from the matching model. Once again, lost voters are generally associated with lower turnout (each missing voter in a block group reduces that neighborhood's turnout by about `r -1*round(bg_model[["coefficients"]][["lost_voters"]]*100, digits = 2)` percentage points), but Models 2 and 4 make clear that this effect is concentrated in Black neighborhoods. In neighborhoods where most residents are Black, each lost voter is associated with a turnout decrease between `r -1*round(bg_model2[["coefficients"]][["lost_voters_black"]]*100, digits = 2)` and `r -1*round(tract_model2[["coefficients"]][["lost_voters_black"]]*100, digits = 2)` percentage points (depending on neighborhood definition).

@@

```{r postest-map, echo = FALSE, message = F, warning = F, fig.width = 5, fig.height = 5, fig.cap = "\\label{fig:postest}Depressive Effect of Felony Disenfranchisement", out.extra = '', fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/dec_block_map.rds")
print(j)
```

@@

During the 2017 mayoral election, felony disenfranchisement laws were responsible for removing an estimated `r comma(nrow(readRDS("./temp/ids_of_lost_voters.rds")))` from New York City neighborhoods. The spatial concentration of these lost voters is striking: as Tables \ref{tab:blockgroup-table} and \ref{tab:tract-table} make clear, these voters were removed from neighborhoods with significantly lower incomes than the rest of the city. They were also removed from neighborhoods that were far less white than the average neighborhood. The systematic removal of voters from these neighborhoods is troubling. However, more than 850,000 votes were cast in the general election of 2017. Although felony disenfranchisement rules have enormous implications for the individuals targeted by them, it is unlikely that the removal of such a small number of voters - concentrated though they are - is likely to have large implications on its own.

Felony disenfranchisement, however, reaches beyond the individuals who are incarcerated. Previous literature has established that felony disenfranchisement likely has impacts on African-American turnout at the state level. This analysis demonstrates that these demobilizing effects intersect with geographical space

>>>>

---
title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City"
author: "Kevin Morris"
date: "April 22, 2019"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
bibliography: "../research/citations.bib"
link-citations: yes
---

@@

Felony disenfranchisement has received much attention over the past few years. This is true in both the academic world (where researchers have dug into the political ramifications in such races as the 2000 presidential election) and in the political world as states such as Florida and Louisiana have gradually moved to dismantle their systems of disenfranchisement. Throughout this process, however, many researchers have focused on the *individual* effects of felony disenfranchisement. They have examined the effect of disenfranchisement on the political participation of individuals, including after they are no longer disenfranchised, but have generally neglected to examine the intersection of spatially concentrated policing and incarceration patterns with the political effects of disenfranchisement. This paper seeks to increase our understanding of the spatial implications of felony disenfranchisement by examining the character of disenfranchisement in New York City.

@@

Of course, despite dire over-incarceration in some neighborhoods, the number of incarcerated individuals is relatively low compared to the number of voters. In New York, for instance, `r comma(readRDS("./temp/number_incarcerated.rds"))` individuals were imprisoned in New York State in early 2019, compared with `r round(readRDS("./temp/active_voters_0319.rds")/1000000, digits = 1)` million actively registered voters. Despite the lower share of residents who are disenfranchised through imprisonment, there is reason to believe the policy impacts more individuals than just those imprisoned. As discussed above, previous research has demonstrated that felony disenfranchisement reduces turnout even among Black voters whose rights are not stripped. This research has found, in particular, that eligible Black voters are less likely to cast a ballot in states where felony disenfranchisement policies are harsher. Understanding the distribution of this effect is important for understanding its political implications. Are voters who live in the same neighborhood as the disenfranchised less likely to cast a ballot because their neighbors have been removed from the community? Or does felony disenfranchisement operate cast its chilling effect on turnout more broadly, crossing even into neighborhoods where voters have not been removed from the community?

@@

The third dataset used to estimate the extent to which a neighborhood is directly impacted by felony disenfranchisement policies in New York State comes from another freedom of information request. The individual-level incarceration data referenced above does not include information about the neighborhood an incarcerated person lived in prior to their conviction. Although the State of New York does not make this information available at the individual level, the author has received prison-admission counts by zip-code for the years XX -- XX.

@@

## Identification of Lost Voters {-}
In this analysis, I offer a slightly different defintion of "lost voter" than much of the literature. Many recent papers have attempted to identify relationships between the number of *potentially* lost voters and turnout. Here, I explore whether the disenfranchisement of voters with a history of participating in elections is related to neighborhood turnout. In this analysis, "lost voters" are individuals ineligible to cast a ballot on a given election day who have cast a ballot in the previous ten years. To study the effect of felony disenfranchisement on voting at the local level (and its potential implications for the distribution of political power at the local level), I use turnout in the most recent election for city-wide office in New York --- the mayoral election which took place on November 7^th^, 2017. Lost voters, therefore, are all individuals who were incarcerated or on parole on November 7^th^, 2017, and had cast a ballot between 2007 and 2016.

@@

After matching neighborhoods, I use a simple regression to test whether turnout in the 2017 mayoral election was different in areas with lost voters. Census tract and block group level turnout rates are calculated using the geocoded voter file. Each voter's record indicates whether the voter participated in the 2017 general election,^[The New York State voter file is not without errors. For instance, the voter file erroneously does not record the author as participating in the 2017 primary election. However, insofar as these errors are not geographically correlated, there is little reason to believe they impact the results of this analysis. Unfortunately, testing the distribution of these errors is not possible.] which are then aggregated to estimate the number of ballots cast in each neighborhood. The number of ballots cast is divided the voting age population.

@@

Matching methodologies, of course, only allow us to test the effect of being treated --- here, losing a voter for the 2017 election. The model above does not allow for different effects on turnout based on *how many* voters a neighborhood lost. In Table \ref{tab:trad-reg} below, I adopt a standard ordinary least squares regression to investigate whether lost voters are associated with lower turnout rates in the 2017 election. This regression uses the same covariates that were used in the matching procedure described above. Each neighborhood is weighted by its voting age population, and robust standard errors are clustered by city council district.

@@

stargazer(bg_model, bg_model2, tract_model, tract_model2,
          column.separate = c(2, 2),
          column.labels = c("Block Group Level", "Tract Level"),
          header = F,
          type = "latex",
          se = list(bg_model_ses, bg_model2_ses, tract_model_ses, tract_model2_ses),
          covariate.labels = c("Lost Voters", "Lost Voters X Share Non-Hispanic Black", "Median Income (Thousands of Dollars)", "Percent Latino", "Percent Non-Hispanic Black", "Percent Non-Hispanic White",
                               "Percent With Some College", "Median Age", "Registration Rate", "Percent Democrats", "Percent Noncitizen", "Percent Won by City Council Representative"),
          dep.var.labels = c("Turnout Rate"),
          title = "(\\#tab:trad-reg) Standard Regression",
          table.placement = "H",
          style = "ajps",
          omit.stat = c("f", "ser"),
          notes = "Robust standard errors (clustered by city council district) in parentheses.",
          omit = "boro")
```
The results presented in Table \ref{tab:trad-reg} align very closely with the estimated effect from the matching model. Once again, lost voters are generally associated with lower turnout (each missing voter in a block group reduces that neighborhood's turnout by about `r -1*round(bg_model[["coefficients"]][["lost_voters"]]*100, digits = 2)` percentage points), but Models 2 and 4 make clear that this effect is concentrated in Black neighborhoods. In neighborhoods where most residents are Black, each lost voter is associated with a turnout decrease of up to `r -1*round(bg_model2[["coefficients"]][["lost_voters_black"]]*100, digits = 2)` percentage points.

@@

```{r postest-map, echo = FALSE, message = F, warning = F, fig.width = 5, fig.height = 5, fig.cap = "\\label{fig:postest}Depressive Effect of Felony Disenfranchisement", out.extra = '', fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/dec_block_map.rds")
j
```

@@

During the 2017 mayoral election, felony disenfranchisement laws were responsible for removing an estimated `r comma(readRDS("./temp/lost_count_nyc.rds"))` from New York City neighborhoods. The spatial concentration of these lost voters is striking: as Tables \ref{tab:blockgroup-table} and \ref{tab:tract-table} make clear, these voters were removed from neighborhoods with significantly lower incomes than the rest of the city. They were also removed from neighborhoods that were far less white than the average neighborhood. The systematic removal of voters from these neighborhoods is troubling. However, more than 850,000 votes were cast in the general election of 2017. Although felony disenfranchisement rules have enormous implications for the individuals targeted by them, it is unlikely that the removal of such a small number of voters --- concentrated though they are --- is likely to have large implications on its own.

Felony disenfranchisement, however, reaches beyond the individuals who are incarcerated. Previous literature has established that felony disenfranchisement likely has impacts on African-American turnout at the state level. This analysis demonstrates that these demobilizing effects intersect with geographical space to systematically depress the vote in neighborhoods where voters are being sent to prison. These communities are already home to some of the poorest and most marginalized voters in the city; they are home to residents who have the least ability to influence policy through other means such as large campaign contributions. The fact that felony disenfranchisement policies appear to be disincentivizing participation on election day --- that these policies are further weakening the ability of these neighborhoods to participate in the democratic process --- is cause for alarm.
